\textbf{Effective application of machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-10-07 submission/manuscript.tex
updates after Pats 2nd final read

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable AMONG individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well non-linear models.

@@

\textbf{Figure S3. Classification performance of ML models across cross
validation when trained on a subset of th dataset.} (A) L2-regularized
logistic regression and (B) Random forest models were trained using the
original study design with 490 subjects and subsets of it with 15, 30,
60, 120, and 245 subjects. The range among the cross-validation AUROC
values within both models at lower sample sizes were much larger than
when the full collection of samples was used to train and validate the
models, but included the ranges observed with the more complete
datasets.

\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

\textbf{Figure S4. Interpretation of the linear ML models with
permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
regression were interpreted using permutation importance using held-out
test set.

@@

\textbf{Figure S5. Training times of ML models when dataset is subsetted
.} (A) L2-regularized logistic regression and (B) Random forest models
were trained using the original study design with 490 subjects and
subsets of it with 15, 30, 60, 120, and 245 subjects. As the size of the
dataset increased, the training times for L2-regularized logistic
regression and random forest models increased linearly.

>>>>

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had higher test AUROC values than the other models
for detecting SRNs (Resampling test from two groups, p \textless{}
0.05). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and
L2-regularized SVM with linear kernel) performed nearly as well
non-linear models. To support our claim, we investigated the
distribution of differences between random forest and L2-regularized
logistic regression AUROC values for each data-split (Figure S3). Our
results underscored that L2-regularized logistic regression perform
nearly as well as random forest.

@@

\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
logistic regression and random forest for each datasplit.} The
percentage of dataplits where the difference between random forest and
L2-regularized logistic regression was higher than or equal to 0 were
0.75, lower than or equal to 0 were 0.25. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

\textbf{Figure S4. Classification performance of ML models across cross
validation when trained on a subset of the dataset.} (A) L2-regularized
logistic regression and (B) Random forest models were trained using the
original study design with 490 subjects and subsets of it with 15, 30,
60, 120, and 245 subjects. The range among the cross-validation AUROC
values within both models at lower sample sizes were much larger than
when the full collection of samples was used to train and validate the
models, but included the ranges observed with the more complete
datasets.

@@

\textbf{Figure S5. Interpretation of the linear ML models with
permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
regression were interpreted using permutation importance using held-out
test set.

\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}

\textbf{Figure S6. Training times of ML models when dataset is subsetted
.} (A) L2-regularized logistic regression and (B) Random forest models
were trained using the original study design with 490 subjects and
subsets of it with 15, 30, 60, 120, and 245 subjects. As the size of the
dataset increased, the training times for L2-regularized logistic
regression and random forest models increased linearly.
